Data Exploration: IUCN and AquaMaps

### 2017-03-15 data_explore/data_explore.Rmd
updating based on WoRMS database.  Turns out the method doesn't improve upon the most recent round of species matches, but it is probably easier for future reference!
updated figures for manuscript

spp_am <- read_csv(file.path(dir_data, 'spp_am.csv')) %>%
  left_join(read_csv(file.path(dir_M, 'git-annex/globalprep/_raw_data/aquamaps/d2015/csv/speciesoccursum.csv')) %>% 
              dplyr::select(am_sid = speciesid, class) %>%
              unique(), 
            by = 'am_sid') %>%
  mutate(paired_map = am_sid %in% spp_map_pairs$am_sid)
### nrow(spp_am %>% filter(paired_map)) ### 2330 spp

iucn_maps_class <- spp_iucn_maps %>% 
  select(-iucn_subpop, -name_verified, -spp_group) %>%
  distinct() %>%
  mutate(class      = str_to_title(class),
         paired_map = iucn_sid %in% spp_map_pairs$iucn_sid)
### nrow(iucn_maps_class %>% filter(paired_map)) ### 2330 spp

taxa_iucn_all <- iucn_maps_class %>%
  filter(!is.na(class)) %>%           ### don't count the unknowns
  mutate(n_iucn_tot = n()) %>%        ### length of whole maps list
  group_by(class) %>%
  summarize(n_iucn_class_all  = n(),
            p_iucn_class_all  = round((n_iucn_class_all/first(n_iucn_tot) * 100), 2))

taxa_iucn_solo <- iucn_maps_class %>%
  filter(!is.na(class)) %>%                 ### don't count the unknowns
  filter(!paired_map) %>%  ### filter list to IUCN only
  mutate(n_iucn_solo = n()) %>%             ### length of IUCN only list
  group_by(class) %>%
  summarize(n_iucn_class_solo = n(),
            p_iucn_class_solo = round((n_iucn_class_solo/first(n_iucn_solo) * 100), 2))

taxa_am_all <- spp_am %>%
  filter(!class %in% c('Not assigned', 'null') & !is.na(class)) %>%
  mutate(n_am_tot = n()) %>%        ### length of whole AM maps list
  group_by(class) %>%
  summarize(n_am_class_all  = n(),
            p_am_class_all  = round((n_am_class_all/first(n_am_tot) * 100), 2))

taxa_am_solo <- spp_am %>%
  filter(!class %in% c('Not assigned', 'null') & !is.na(class)) %>%
  filter(!paired_map) %>%  ### filter list to AM only
  mutate(n_am_solo = n()) %>%       ### length of AM only list
  group_by(class) %>%
  summarize(n_am_class_solo = n(),
            p_am_class_solo = round((n_am_class_solo/first(n_am_solo) * 100), 2))

taxa_both <- spp_am %>%
  filter(paired_map) %>%
  filter(!class %in% c('Not assigned', 'null')) %>%
  mutate(n_both_tot = n()) %>%
  group_by(class) %>%
  summarize(n_class_both = n(),
            p_class_both = round(n_class_both/first(n_both_tot) * 100, 2))

taxa_df <- taxa_am_all %>%
  full_join(taxa_am_solo,   by = 'class') %>%
  full_join(taxa_iucn_all,  by = 'class') %>%
  full_join(taxa_iucn_solo, by = 'class') %>%
  full_join(taxa_both,      by = 'class') %>%
  filter(!(is.na(n_iucn_class_all) & p_am_class_all < 0.5))

taxa_omitted <- taxa_am_all %>%
  full_join(taxa_am_solo,   by = 'class') %>%
  full_join(taxa_iucn_all,  by = 'class') %>%
  full_join(taxa_iucn_solo, by = 'class') %>%
  full_join(taxa_both,      by = 'class') %>%
  filter(is.na(n_iucn_class_all) & p_am_class_all < 0.5) %>%
  left_join(read_csv(file.path(dir_data, 'sciname_lookups/scinames.csv')) %>%
              select(class, kingdom, phylum) %>%
              mutate(class = str_to_title(class)),
            by = 'class') %>%
  mutate(class = paste('other', str_to_title(kingdom), sep = ' ')) %>%
  select(-kingdom, -phylum) %>%
  distinct()

# knitr::kable(taxa_df %>%
#                dplyr::select(class, 
#                              AquaMaps = n_am_class_all, 
#                              IUCN = n_iucn_class_all, 
#                              Both = n_class_both))

write_csv(taxa_df, file.path(dir_data, 'taxa_df.csv'))
write_csv(taxa_omitted, file.path(dir_data, 'taxa_omitted.csv'))

@@

taxa_df <- read_csv(file.path(dir_data, 'taxa_df.csv'))
taxa_omitted <- read_csv(file.path(dir_data, 'taxa_omitted.csv'))

n_spp_omitted <- taxa_omitted$n_am_class_all %>% sum()

@@

taxa_all_df <- taxa_df1 %>%
  group_by(dataset) %>%
  summarize(class = 'Total',
            AquaMaps_total = sum(AquaMaps)  + n_spp_omitted + sum(Paired),
            IUCN_total = sum(IUCN) + sum(Paired),
            n_total  = sum(n_total) + n_spp_omitted,
            IUCN_end = 100 * IUCN_total/n_total,
            IUCN_start = 0,
            AquaMaps_end   = 100,
            AquaMaps_start = 100 * (n_total - AquaMaps_total)/n_total,
            Paired_end = AquaMaps_start,
            Paired_start = IUCN_end) %>%
  select(-dataset) %>% distinct() %>%
  gather(key = dataset,  value = start, ends_with('start')) %>%
  gather(key = dataset2, value = end, ends_with('end')) %>%
  mutate(dataset = str_replace(dataset, '_start', ''),
         dataset2 = str_replace(dataset2, '_end', '')) %>%
  filter(dataset == dataset2) %>%
  select(-dataset2, -AquaMaps_total, -IUCN_total)

taxa_df2 <- taxa_all_df %>%
  bind_rows(taxa_all_df %>%
              mutate(class = '',
                     start = 0,
                     end   = 0,
                     n_total = 0)) %>%
  bind_rows(taxa_df1 %>%
              select(-IUCN, -AquaMaps, -Paired) %>%
              arrange(desc(n_total))) %>%
  bind_rows(taxa_omitted %>%
              group_by(class) %>%
              summarize(n_total = sum(n_am_class_all),
                        dataset = 'AquaMaps',
                        start = 0,
                        end = 100)) %>%
  mutate(class = factor(class, levels = unique(.$class), ordered = TRUE),
         n_label = ifelse(n_total > 0, paste0('n = ', n_total), ''),
         dataset = factor(dataset, levels = c('AquaMaps', 'Paired', 'IUCN'), ordered = TRUE))

@@

>>>>

spp_match_key <- spp_map_pairs %>%
  select(iucn_sid, am_sid)

am_maps_class <- read_csv(file.path(dir_data, 'spp_am.csv')) %>%
  select(am_sid, sciname) %>%
  left_join(read_csv(file.path(dir_M, 'git-annex/globalprep/_raw_data/aquamaps/d2015/csv/speciesoccursum.csv')) %>% 
              dplyr::select(am_sid = speciesid, am_class = class) %>%
              unique(), 
            by = 'am_sid') %>%
  left_join(spp_match_key, by = 'am_sid')
### nrow(spp_am %>% filter(paired_map)) ### 2330 spp

iucn_maps_class <- spp_iucn_maps %>% 
  select(iucn_sid, sciname_iucn = sciname, iucn_class = class) %>%
  distinct() %>%
  mutate(iucn_class = str_to_title(iucn_class))
### nrow(iucn_maps_class %>% filter(paired_map)) ### 2330 spp

spp_class_all <- am_maps_class %>%
  full_join(iucn_maps_class, by = 'iucn_sid') %>%
  mutate(am_class = ifelse(am_class %in% c('null', 'not assigned'), NA, am_class),
         class    = if_else(is.na(am_class), iucn_class, am_class))

### Fix some missing class values
# spp_class_all1 %>% filter(is.na(class))
#       am_sid                   sciname am_class iucn_sid sciname_iucn iucn_class class
# 1  Fis-22442 Metapenaeopsis tarawensis     <NA>       NA         <NA>       <NA>  <NA> # Malacostraca
# 2 Fis-170350          Oenopota eriopis     <NA>       NA         <NA>       <NA>  <NA> # Gastropoda
# 3 Fis-170316          Yarica hyalosoma     <NA>       NA         <NA>       <NA>  <NA> # Actinopterygii
spp_class_all <- spp_class_all %>%
  mutate(class = case_when(
    is.na(.$class) & .$am_sid == 'Fis-22442' ~ 'Malacostraca',
    is.na(.$class) & .$am_sid == 'Fis-170350' ~ 'Gastropoda',
    is.na(.$class) & .$am_sid == 'Fis-170316' ~ 'Actinopterygii',
    TRUE ~ .$class))

write_csv(spp_class_all, file.path(dir_data, 'spp_class_all.csv'))

``` {r break down by taxa and data source}

spp_class_all <- read_csv(file.path(dir_data, 'spp_class_all.csv'))

taxa_iucn_all <- spp_class_all %>%
  filter(!is.na(iucn_sid)) %>%
  mutate(n_iucn_tot = n()) %>%        ### length of whole maps list
  group_by(class) %>%
  summarize(n_iucn_class_all  = n(),
            p_iucn_class_all  = round((n_iucn_class_all/first(n_iucn_tot) * 100), 2))

taxa_iucn_solo <- spp_class_all %>%
  filter(!is.na(iucn_sid) & is.na(am_sid)) %>%  ### filter list to IUCN only
  mutate(n_iucn_solo = n()) %>%             ### length of IUCN only list
  group_by(class) %>%
  summarize(n_iucn_class_solo = n(),
            p_iucn_class_solo = round((n_iucn_class_solo/first(n_iucn_solo) * 100), 2))

taxa_am_all <- spp_class_all %>%
  filter(!is.na(am_sid)) %>%
  mutate(n_am_tot = n()) %>%        ### length of whole AM maps list
  group_by(class) %>%
  summarize(n_am_class_all  = n(),
            p_am_class_all  = round((n_am_class_all/first(n_am_tot) * 100), 2))

taxa_am_solo <- spp_class_all %>%
  filter(!is.na(am_sid) & is.na(iucn_sid)) %>%
  mutate(n_am_solo = n()) %>%       ### length of AM only list
  group_by(class) %>%
  summarize(n_am_class_solo = n(),
            p_am_class_solo = round((n_am_class_solo/first(n_am_solo) * 100), 2))

taxa_both <- spp_class_all %>%
  filter(!is.na(am_sid) & !is.na(iucn_sid)) %>%
  mutate(n_both_tot = n()) %>%
  group_by(class) %>%
  summarize(n_class_both = n(),
            p_class_both = round(n_class_both/first(n_both_tot) * 100, 2))

taxa_all <- taxa_am_all %>%
  full_join(taxa_am_solo,   by = 'class') %>%
  full_join(taxa_iucn_all,  by = 'class') %>%
  full_join(taxa_iucn_solo, by = 'class') %>%
  full_join(taxa_both,      by = 'class') %>%
  mutate(small_class = (is.na(n_iucn_class_all) & p_am_class_all < 0.5))

taxa_df <- taxa_all %>%
  filter(!small_class) %>%
  select(-small_class)

class_to_kingdom <- read_csv(file.path(dir_M, 'git-annex/globalprep/_raw_data',
                                       'aquamaps/d2015/csv/speciesoccursum.csv')) %>%
  select(class, kingdom, phylum) %>%
  mutate(class = str_to_title(class)) %>%
  distinct()

taxa_other <- taxa_all %>%
  filter(small_class) %>%
  select(-small_class) %>%
  left_join(class_to_kingdom,
            by = 'class') %>%
  mutate(class = paste('other', str_to_title(kingdom), sep = ' ')) %>%
  select(-kingdom, -phylum) %>%
  group_by(class) %>%
  summarize(n_am_class_all = sum(n_am_class_all),
            p_am_class_all = sum(p_am_class_all),
            n_am_class_solo = sum(n_am_class_solo),
            p_am_class_solo = sum(p_am_class_solo))

write_csv(taxa_df, file.path(dir_data, 'taxa_df.csv'))
write_csv(taxa_other, file.path(dir_data, 'taxa_other.csv'))

@@

taxa_df <- read_csv(file.path(dir_data, 'taxa_df.csv'))
taxa_other <- read_csv(file.path(dir_data, 'taxa_other.csv'))

n_spp_other <- taxa_other$n_am_class_all %>% sum()

@@

taxa_total_df <- taxa_df1 %>%
  group_by(dataset) %>%
  summarize(class = 'Total',
            AquaMaps_total = sum(AquaMaps)  + n_spp_other + sum(Paired),
            IUCN_total = sum(IUCN) + sum(Paired),
            n_total  = sum(AquaMaps) + sum(IUCN) + n_spp_other + sum(Paired),
            IUCN_end = 100 * IUCN_total/n_total,
            IUCN_start = 0,
            AquaMaps_end   = 100,
            AquaMaps_start = 100 * (n_total - AquaMaps_total)/n_total,
            Paired_end = AquaMaps_start,
            Paired_start = IUCN_end) %>%
  select(-dataset) %>% distinct() %>%
  gather(key = dataset,  value = start, ends_with('start')) %>%
  gather(key = dataset2, value = end, ends_with('end')) %>%
  mutate(dataset = str_replace(dataset, '_start', ''),
         dataset2 = str_replace(dataset2, '_end', '')) %>%
  filter(dataset == dataset2) %>%
  select(-dataset2, -AquaMaps_total, -IUCN_total)

taxa_df2 <- taxa_total_df %>%
  bind_rows(taxa_total_df %>%
              mutate(class = '',
                     start = 0,
                     end   = 0,
                     n_total = 0)) %>%
  bind_rows(taxa_df1 %>%
              select(-IUCN, -AquaMaps, -Paired) %>%
              arrange(desc(n_total))) %>%
  bind_rows(taxa_other %>%
              group_by(class) %>%
              summarize(n_total = sum(n_am_class_all),
                        dataset = 'AquaMaps',
                        start = 0,
                        end = 100)) %>%
  mutate(class = factor(class, levels = unique(.$class), ordered = TRUE),
         n_label = ifelse(n_total > 0, paste0('n = ', n_total), ''),
         dataset = factor(dataset, levels = c('AquaMaps', 'Paired', 'IUCN'), ordered = TRUE))

@@

spp_list_quads <- read_csv(file.path(dir_data, 'spp_list_quads.csv'))
